---
title: "Snow in the Last PMIP-3 Last Millennium "
output: html_notebook
---

# Setup

```{r setup}
library(raster)
library(stars)
#install.packages('../tidyEOF_0.1.0.tar.gz', repos = NULL, type = 'source')
library(tidyEOF)
library(tidyverse)
library(scico)
library(patchwork)

bbox <- extent(c(-125, -102, 31, 49))

state_names <- c('arizona', 'new mexico', 'colorado', 
                 'california', 'utah', 'nevada', 
                 'oregon', 'washington', 'idaho',
                 'wyoming', 'montana')

states_wus <- maps::map('state', regions = state_names, 
                    fill = TRUE, plot = FALSE) %>% 
  st_as_sf()

load('../tidyEOF/R/sysdata.rda')
```

## Data import and preproceesing

Write a function to preprocess the data.

```{r}
preprocess <- function(x, var, bbox, flip = FALSE, regrid = FALSE, daily = FALSE) {
  maps <- brick(x, varname = var)

  indices <- getZ(maps) %>%
    str_detect('-03-')%>%
    which()

  raster::subset(maps, indices) %>%
    {if(flip) rotate(.) else .} %>%
    raster::crop(bbox, snap = 'out') %>% # snap = out for coarser grids
    {if(daily) mean(.) else .} %>% # if a file is one year of daily values
    {if(regrid) aggregate(., fact = 2) else .} # resample to lower res to speed up analysis
}
```



Read in and preprocess the ensemble. The MRI-CGCM3 simulation is in two parts because its resolution is so high. Read it in first, combine the two layers, and use as a reference to resample the other ensemble members.

```{r}
mri <- c('data/snw_LImon_MRI-CGCM3_past1000_r1i1p1_085001-134912.nc',
  'data/snw_LImon_MRI-CGCM3_past1000_r1i1p1_135001-184912.nc') %>%
  map(preprocess, var = 'snw', bbox, flip = TRUE) %>%
  brick() %>%
  resample(as(cera, 'Raster')) %>%
  clamp(lower = 0) %>%
  mask(., all(near(., 0)), maskvalue = 1)

rasts <- c('data/snw_LImon_MIROC-ESM_past1000_r1i1p1_085001-184912.nc',
  'data/snw_LImon_HadCM3_past1000_r1i1p1_085001-185012.nc',
  'data/snw_LImon_MPI-ESM-P_past1000_r1i1p1_085001-184912.nc',
  'data/snw_LImon_CCSM4_past1000_r1i1p1_085001-185012.nc') %>%
  map(preprocess, var = 'snw', bbox, flip = TRUE) %>%
  map(resample,as(cera, 'Raster')) %>%
  map(clamp, lower = 0) %>% # the resampling procedure introduces negative values, this sets them to zero
  map(~mask(., all(near(., 0)), maskvalue = 1)) #%>% # mask cells that never experience snow
  #map(mask, mri[[1]])

rasts[[5]] <- mri

rast_list <- map(rasts, ~st_as_stars(.) %>%
               .[,,,1:1000] %>% # hadcm3 and ccsm4 have one extra year
  setNames('swe') %>%
  st_set_dimensions('band', names = 'time', values = 850:1849))

gcm_ensemble <- rast_list %>%
    append(list(along = list(model = c('MIROC-ESM', 'HadCM3', 'MPI-ESM-P', 'CCSM4', 'MRI-CGCM3')))) %>%
do.call(c, .) %>%
  mutate(swe = units::set_units(swe, mm)) %>%
  st_crop(st_as_sf(cera[,,,1]))
```


```{r}
ggplot() +
  geom_stars(data = st_apply(gcm_ensemble, c('x', 'y', 'model'), mean)) +
  facet_wrap(~model) +
  scale_fill_distiller(na.value = NA) +
  geom_sf(data = states_wus, fill = NA) +
  theme_bw()
```


```{r}
patterns <- map(1:5, ~get_patterns(slice(gcm_ensemble, 'model', .)), k = 4)
map(patterns, plot_eofs)

map(1:5, ~ ggplot() +
  geom_stars(data = get_correlation(slice(gcm_ensemble, 'model', .), patterns[[.]])) +
        geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~PC, nrow = 1) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation', na.value = NA) +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void())

```


## Reconstruciton
-- start here, make sure gcm ensembles same format and cells as cera
```{r}
recon_test1 <- predict_cca(get_patterns(filter(cera, time >= 1982), 10), 
            get_patterns(filter(prism, time <= 2010), 9), 
            slice(gcm_ensemble, 'model', 4)[,,,1:700],
            k = 7)

recon_test2 <- predict_cca(get_patterns(filter(cera, time >= 1982), 10), 
            get_patterns(filter(prism, time <= 2010), 9), 
            slice(gcm_ensemble, 'model', 4)[,,,701:1000],
            k = 7)

recon_test <- c(recon_test1, recon_test2)
```

```{r}
plot(st_apply(recon_test, 1:2, mean));plot(st_apply(prism, 1:2, mean))
plot(st_apply(recon_test, 1:2, sd));plot(st_apply(prism, 1:2, sd))
```

```{r}
plot(gcm_ensemble[,,,1,4] %>% st_crop(st_as_sf(cera[,,,1])))
plot(gcm_ensemble[,,,1,4])
plot(cera[,,,1])
plot(ccsm[,,,1])
```


## Lattice Krig
```{r}
resids <- prism_recon %>%
  left_join(prism_dat, by = c('x', 'y', 'year'), suffix = c('_recon', '_obs')) %>%
  group_by(x, y) %>%
  mutate(error = SWE_recon - SWE_obs,
           error_sd = error / sd(SWE_obs), # pearson residuals
           error_sd2 = error / sd(error), # standardized (internally studentized)
         snowy = median(SWE_obs) > 3)
```

```{r}
out1$rho.MLE
out2$rho.MLE
sqrt(out1$rho.MLE.FULL)
sqrt(out2$rho.MLE.FULL)
```


```{r}
# these simulate with marginal unit variance, need to multiply by sill to recover original
gSim <- LKrig.sim(st_coordinates(pts), out1$LKinfo, M = 6) * sqrt(out1$rho.MLE.FULL)

gSim2 <- LKrig.sim(st_coordinates(pts), out2$LKinfo, M = 6) * sqrt(out2$rho.MLE.FULL)
```

```{r}
resids %>%
  group_by(x,y) %>%
  summarise(swe_sd = sd(error)) %>%
  bind_cols(as_tibble(gSim)) %>%
  pivot_longer(V1:V6) %>%
    ggplot(aes(x, y)) +
  geom_raster(aes(fill = value * swe_sd)) +
  facet_wrap(~name) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-300, 300)) +
  coord_quickmap() +
  theme_void()

resids %>%
  filter(year == 1982) %>%
  bind_cols(as_tibble(gSim)) %>%
  pivot_longer(V1:V6) %>%
    ggplot(aes(x, y)) +
  geom_raster(aes(fill = value)) +
  facet_wrap(~name) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-3, 3)) +
  coord_quickmap() +
  theme_void()

resids %>%
  group_by(x,y) %>%
  summarise(swe_sd = sd(error)) %>%
  bind_cols(as_tibble(gSim2)) %>%
  pivot_longer(V1:V6) %>%
    ggplot(aes(x, y)) +
  geom_raster(aes(fill = value * swe_sd)) +
  facet_wrap(~name) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-300, 300)) +
  coord_quickmap() +
  theme_void()

resids %>%
  filter(year == 1982) %>%
  bind_cols(as_tibble(gSim2)) %>%
  pivot_longer(V1:V6) %>%
    ggplot(aes(x, y)) +
  geom_raster(aes(fill = value)) +
  facet_wrap(~name) +
  #scale_fill_distiller(palette = 'RdBu', limits = c(-3, 3)) +
  coord_quickmap() +
  theme_void()
```

```{r}
pts <- resids %>%
  filter(year == 1982) %>%
  st_as_sf(coords = c('x','y'), crs = 4326)

dat <- resids %>%
  pivot_wider(id_cols = c(x,y), names_from = year, values_from = error_sd2)
dat2 <- resids %>%
  pivot_wider(id_cols = c(x,y), names_from = year, values_from = error)
```

interesting, lattice kriging the standardized and unstandardized doesn't seem to make a difference. 
```{r}
out1<- LatticeKrig(st_coordinates(pts), dat[,3:38])
out2<- LatticeKrig(st_coordinates(pts), dat2[,3:38])
```

```{r}
out1
out2
```
## other stuff


```{r}
get_corrs2 <- function(dat, patterns) { # should be able to take arbitrary number of amps rather than first 4
  amps <- patterns$amplitudes %>%
    spread(PC, amplitude, sep ='')

  dat %>%
    group_by(x, y) %>%
    nest %>%
    # can use corrr here 
    mutate(corrs = map(data,  ~inner_join(., amps, by = 'year') %>%
                         summarise(PC1 = cor(value, PC1),
                                   PC2 = cor(value, PC2),
                                   PC3 = cor(value, PC3),
                                   PC4 = cor(value, PC4)))) %>%
    select(-data) %>%
    unnest(corrs) %>%
    ungroup()
}

corrs <- dat %>%
  map( ~rename(., value = SWE)) %>%
  map2(patterns, get_corrs2) %>%
  setNames(c("MIROC-ESM", "HadCM3", "MPI-ESM-P", "CCSM4", "MRI-CGCM3"))

corrs %>%
  bind_rows(.id = 'model') %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
  #      geom_sf(data = states_wus, fill = NA, color = 'black', alpha = .5) +
  facet_grid(PC~model) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void() +
  coord_quickmap()
```

```{r}
patterns %>%
  map(~.$amplitudes) %>%
  setNames(c("MIROC-ESM", "HadCM3", "MPI-ESM-P", "CCSM4", "MRI-CGCM3")) %>%
  bind_rows(.id = 'model') %>%
ggplot(aes(year, amplitude)) + geom_line() + facet_grid(PC~model)+ geom_smooth()
  
 patterns[[1]]$amplitudes %>%
  group_by(PC) %>%
  filter(amplitude == max(amplitude) | amplitude == min(amplitude)) %>%
   mutate(sign = sign(amplitude)) %>%
   left_join(dat[[1]]) %>%
  ggplot(aes(x, y, fill = SWE)) +
  geom_raster() +
  facet_grid(PC~sign) +
  coord_quickmap() +
  scale_fill_viridis_c() +
  theme_void()

#reof_obs$amps <- reof_obs$amps %>%
#  mutate(reof_amp = if_else(PC == 'PC6', reof_amp * -1, reof_amp))
```


```{r}
corrs[[1]]$PC1 %*% corrs[[1]]$PC1

corrs[[1]]$PC1 %*% -corrs[[1]]$PC1
cor(corrs[[1]]$PC1, corrs[[5]]$PC1)

```
Try the same thing across time
```{r}
test <- dat[[4]] %>%
  mutate(century = floor(year/100)) %>%
  group_nest(century) %>%
  mutate(patterns = map(data, get_patterns, k = 4),
         data = map(data, ~rename(., value = SWE)),
        corrs = map2(data, patterns, get_corrs2))

test %>%
  select(century, corrs) %>%
  unnest(corrs) %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
    #    geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_grid(PC~century) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void() +
  coord_quickmap()
```

```{r}
test_fun <- function(x) {
  x %>% mutate(century = floor(year/100)) %>%
  group_nest(century) %>%
  mutate(data = map(data, ~group_by(., x,y) %>% filter(all(SWE > 0)) %>% ungroup),
         patterns = map(data, get_patterns, k = 4),
         data = map(data, ~rename(., value = SWE)),
        corrs = map2(data, patterns, get_corrs2))
}
  
tmp <- map(dat, test_fun)
########3 start here, I've done the one functio right, now try the matching!
tmp[[1]] %>%
  select(century, corrs) %>%
  unnest(corrs) %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
    #    geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_grid(PC~century) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void() +
  coord_quickmap()
```


```{r}
library(corrr)
test %>%
  select(century, corrs) %>%
  unnest(corrs)

  
  
newfun <- function(x) {
    x %>%
      as.data.frame() %>%
  rownames_to_column() %>%
  pivot_longer(PC1:PC4) %>%
  group_by(rowname) %>%
  filter(abs(value) == max(abs(value))) %>%
    ungroup
}


key <- test %>%
  mutate(cor2 = map(corrs, ~cor(select(test$corrs[[1]], -x, -y), select(., -x, -y))),
         test = map(cor2, newfun)) %>%
  select(century, test) %>%
  unnest(test)

newfun_2 <-  function(x) {
  x %>%
    mutate(cor2 = map(corrs, ~cor(select(x$corrs[[1]], -x, -y), select(., -x, -y))),
         test = map(cor2, newfun)) %>%
  select(century, test) %>%
  unnest(test)
}
#key isn't working  for some
key2 <- map(tmp[[1]], newfun_2)

t2 <- tmp[[1]] %>%
    mutate(cor2 = map(corrs, ~cor(select(tmp[[1]]$corrs[[1]], -x, -y), select(., -x, -y))))#,
         cors = map(cor2, newfun)) %>%
  select(century, cors) %>%
  unnest(cors)
key

t2$cor2

cor(select(tmp[[1]]$corrs[[1]], -x, -y),select(tmp[[1]]$corrs[[1]], -x, -y))
tmp[[1]]$corrs[[1]]
# now join this back to the plot or something?
```
why are there replicatns
```{r}
test %>%
  select(century, corrs) %>%
  unnest(corrs) %>%
  gather(PC, value, PC1:PC4) %>%
  left_join(key, by = c('century', 'PC' = 'rowname')) %>%
  mutate(value.x = if_else(sign(value.y) == 1, value.x * -1, value.x)) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value.x)) +
    #    geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_grid(name~century) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void() +
  coord_quickmap()
```

```{r}
tmp[[1]] %>%
  select(century, corrs) %>%
  unnest(corrs) %>%
  gather(PC, value, PC1:PC4) %>%
  left_join(key2[[1]], by = c('century', 'PC' = 'rowname')) %>%
  mutate(value.x = if_else(sign(value.y) == 1, value.x * -1, value.x)) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value.x)) +
    #    geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_grid(name~century) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void() +
  coord_quickmap()
```


```{r}
 dat %>%
  walk(~get_anomalies(.) %>%
  get_pcs() %>%
  get_eigenvalues() %>% 
  plot_scree(k = 6, kmax = 10) +
  scale_color_brewer(palette = 'Spectral'))

plot_sc
```


to do teleconnections, need the sst layers from the models



```{r}
miroc <- 'data/snw_LImon_MIROC-ESM_past1000_r1i1p1_085001-184912.nc' %>%
  preprocess('snw', bbox, flip = TRUE) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)

miroc_patterns <- get_patterns(miroc, k = 4)

swe_corr_miroc <- miroc %>%
  rename(value = SWE) %>%
  get_corrs(miroc_patterns)

swe_corr_miroc %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
        geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~PC, nrow = 1) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void()
```

```{r}
hadcm3 <- 'data/snw_LImon_HadCM3_past1000_r1i1p1_085001-185012.nc' %>%
  preprocess('snw', bbox, flip = TRUE) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)

hadcm3_patterns <- get_patterns(hadcm3, k = 4)

swe_corr_hadcm3 <- hadcm3 %>%
  rename(value = SWE) %>%
  get_corrs(hadcm3_patterns)

swe_corr_hadcm3 %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
        geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~PC, nrow = 1) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void()
```

```{r}
mpi <- 'data/snw_LImon_MPI-ESM-P_past1000_r1i1p1_085001-184912.nc' %>%
  preprocess('snw', bbox, flip = TRUE) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)

mpi_patterns <- get_patterns(mpi, k = 4)

swe_corr_mpi <- mpi %>%
  rename(value = SWE) %>%
  get_corrs(mpi_patterns)

swe_corr_mpi %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
        geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~PC, nrow = 1) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void()
```

```{r}
ccsm <- 'data/snw_LImon_CCSM4_past1000_r1i1p1_085001-185012.nc' %>%
  preprocess('snw', bbox, flip = TRUE) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)

ccsm_patterns <- get_patterns(ccsm, k = 4)

swe_corr_ccsm <- ccsm %>%
  rename(value = SWE) %>%
  get_corrs(ccsm_patterns)

swe_corr_ccsm %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
        geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~PC, nrow = 1) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void()
```
```{r}
mri1 <- 'data/snw_LImon_MRI-CGCM3_past1000_r1i1p1_085001-134912.nc' %>%
  preprocess('snw', bbox, flip = TRUE) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)
mri2 <- 'data/snw_LImon_MRI-CGCM3_past1000_r1i1p1_135001-184912.nc' %>%
  preprocess('snw', bbox, flip = TRUE) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)

mri <- rbind(mri1, mri2)

mri_patterns <- get_patterns(mri, k = 4)

swe_corr_mri <- mri %>%
  rename(value = SWE) %>%
  get_corrs(mri_patterns)

swe_corr_mri %>%
  gather(PC, value, PC1:PC4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = value)) +
        geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~PC, nrow = 1) +
scale_fill_scico(palette = 'broc', limits = c(-1, 1), name = 'Correlation') +
    labs(x = 'Longitude', y = 'Latitude') +
  theme_void()
```
